Writing Trash Bin

### 2024-07-25 paper.qmd
Made some final changes to ""visualising uncertainty as signal" and started "transforming space" section

<!-- TODO 
- convert to American spelling
- May not need to reference the same paper again and again
- I can do a more formal proof of the decision making thing using decision science. I left it out but it can be added if that section needs to be more thorough.
- fix the colour value map
-->

@@

@fig-choropleth depicts a choropleth map of the counties of Iowa. Each of these counties are colored according to an estimate of average daily temperature that was simulated so that the values followed a clear spatial trend (hot in the middle of the map, and cold on the outside). The variance of these estimates were simulated such that a hypothesis test would indicate the existence of a spatial trend in the low variance map, while the trend in the high variance map should be indistinguishable from noise. Is this distinction in validity of the spatial trend clear in in the map? Is the validity of the trend communicated through the visualisation?

@@

@fig-bivariate a variation of the choropleth map, where we have a two dimensional color palette. Not only is temperature mapped to hue, but variance is also mapped to saturation. While these two maps *do* look visually different (which was not the case in the choropleth map) the spatial trend is still clearly visible in both graphics. This means the uncertainty *is* being communicated, however the primary take away in the graphic is the spatial trend (that does not exist). The graphic no longer has a transparency issue as the uncertainty *is* technically included, but is this a good uncertainty visualisation? And if not, why?

@@

p2a1
p2b2
show_pal2(colours = bivariatepal2, ncol=8, borders=NA, myxlab = "Temperature", myylab = "Variance", breaks = 21:29, breaks2 = 0:4)

@@

transpal <- as.vector(matrix(unlist(x), ncol = 4, byrow = TRUE))

@@

# Visualising uncertainty and signal in a new space
Instead of hoping that uncertainty might collapse signal values into a single dimension, we can do some of that work ourselves.

by using a VSUP which collapses the colour space such that high uncertainty values cannot be extracted. It is unclear how useful VSUPs are at actually combining signal and noise and therefore suppressing plot level insights, as they have only been tested on simple value extraction tasks that require evaluating a single point [@Correll2018; @Ndlovu2023] rather than looking for spatial relationships (which is arguably what maps are for). Following along with this trend, the next way we might consider visualising uncertainty is to combine uncertainty and signal at the earlier stage so the "suppressed signal" is represented by a single variable. This statistic can then be expressed in a one dimensional colour space, which is a method adopted by the Bayesian surprise metric map [@Ndlovu2023] and the exceedance probability map [Lucchesi2017].

This problem has been noticed by others in the field, that typically use this method (specifically in the spatial uncertainty context) and a desire for representations that integrate uncertainty and signal is one of the reasons for the invention of the value-suppressing uncertainty pallet [@Correll2018].

Value-suppressing uncertainty pallets (VSUP) were developed as a method that would allow the signal and the noise to be interpreted together such that insights gained by the viewers of a plot are appropriately *suppressed* by the uncertainty. Hence the name of the pallet. depicted this map colouring approach and several other extensions on the typical choropleth map that differ in where in the visualisation process they combine the noise and signal information into a single concept of "valid signal".

### Value Supressing Uncertainty Palettes

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-vsup
#| fig-cap: "A map made with a VSUP. The counties of Iowa are coloured acording to it's average daily temperature and the variance in temperature, although the space they have been mapped to is slightly warped."
#| fig-subcap: 
#|   - "High Variance Data"
#|   - "Low Variance Data"
#|   - "VSUP Palette"
#| layout-ncol: 3
#| layout-valign: "bottom"

@@

### Uncertainty and inference
The ramifications of this is seen throughout the literature, as a survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a]. Some works [@Hullman2018; @Maceachren2012; @Thomson2005] get around this definition issue by focusing narrowly on specific terms defined mathematically, such as probability, variance, error, or precision. Others [@Griethe2006; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values, but ignore these concepts when translating that uncertainty to a visualisation. Existing mathematical definition of uncertainty (or related concepts) have been unable to unify this range of diverse concepts. For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, and information theory tries to quantify uncertainty using the idea of entropy, but the disconnect between the broad concept of uncertainty and what we can reliably quantify remains a problem.

>>>>

<!-- TODO 
- convert to American spelling
- May not need to reference the same paper again and again
- examples that aren't maps? Sure are a lot of maps.
-->

@@

## The choropleth map
@fig-choropleth depicts a choropleth map of the counties of Iowa. Each of these counties are colored according to an estimate of average daily temperature that was simulated so that the values followed a clear spatial trend (hot in the middle of the map, and cold on the outside). The variance of these estimates were simulated such that a hypothesis test would indicate the existence of a spatial trend in the low variance map, while the trend in the high variance map should be indistinguishable from noise. Is this distinction in validity of the spatial trend clear in in the map? Is the validity of the trend communicated through the visualisation?

@@

## The bivariate map
@fig-bivariate a variation of the choropleth map, where we have a two dimensional color palette. Not only is temperature mapped to hue, but variance is also mapped to saturation. While these two maps *do* look visually different (which was not the case in the choropleth map) the spatial trend is still clearly visible in both graphics. This means the uncertainty *is* being communicated, however the primary take away in the graphic is the spatial trend (that does not exist). The graphic no longer has a transparency issue as the uncertainty *is* technically included, but is this a good uncertainty visualisation? And if not, why?

@@

#p2a1
#p2b2
#show_pal2(colours = bivariatepal2, ncol=8, borders=NA, myxlab = "Temperature", myylab = "Variance", breaks = 21:29, breaks2 = 0:4)

@@

transpal <- as.vector(matrix(unlist(transpal), ncol = 4, byrow = TRUE))

@@

# Visualising uncertainty and signal in a new space
Instead of hoping that uncertainty might collapse signal values into a single dimension, we can do some of that work ourselves, and uncertainty visualisation authors already have.

### Value Supressing Uncertainty Palettes
The Value Suppressing Uncertainty Palette (VSUP) [@Correll2018], was designed with the intention of preventing high uncertainty values from being extracted from a map. Since the palette was designed with the extraction of individual values in mind and it has only been tested on simple value extraction tasks [@Correll2018] or search tasks [@Ndlovu2023], it is unclear how effective the method is at suppressing broader insights such as spatial trends.

@fig-vsup is a visualisation of the Iowa temperature data using a VSUP to color the counties. The low uncertainty case still has a visible spatial trend, while the spatial trend in the high uncertainty map has functionally disappeared. This means the VSUP has successfully performed signal suppression for our data.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-vsup
#| fig-cap: "A map made with a VSUP. The counties of Iowa are coloured acording to it's average daily temperature and the variance in temperature, although the space they have been mapped to is slightly warped. Similar to the bivariate map, temperature is mapped to hue while variance is mapped to saturation. This map successfully reduces the visibility of the spatial trend in the high uncertainty case while maintaining the visibility of the spatial trend in the low uncertainty case."
#| fig-subcap: 
#|   - "High Variance Data"
#|   - "Low Variance Data"
#|   - "VSUP Palette"
#| layout-ncol: 3
#| layout-valign: "bottom"

@@

Now that we have successfully suppressed our invalid signal, we must return to the original criteria we had for signal supression and ask ourselves if we meet all of them. We want to make sure that in our efforts to suppress invalid signals, we have not also suppressed statistically *valid* signals in the process.

## Can we have uncertinty visualisations for EDA?
A slight problem with uncertainty visualisation is that the uncertainty and the purpose of visualisation are somewhat at odds with one another. This problem becomes much more explicit through the use of the VSUP.

### What is uncertainty?
There does not seem to be a clear agreed upon definition of uncertainty which would obviously create some issues when we want to visualize it. You cant measure something you cant define, you cant quantify something you cant measure, and you cant visualise something you cant quantify.

The ramifications of this problem is seen repeatedly in the uncertaint visualisation literature. A survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a]. Some works [@Hullman2018; @Maceachren2012; @Thomson2005] ignore the problem get around this definition issue by focusing narrowly on specific terms defined mathematically, such as probability, variance, error, or precision. Others [@Griethe2006; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values, but ignore these concepts when translating that uncertainty to a visualisation. Existing mathematical definition of uncertainty (or related concepts) have been unable to unify this range of diverse concepts. For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, and information theory tries to quantify uncertainty using the idea of entropy, but the disconnect between the broad concept of uncertainty and what we can reliably quantify remains a problem.

### 2024-07-25 writingdumpster.qmd
Made some final changes to ""visualising uncertainty as signal" and started "transforming space" section

# Removed Sections
@Kinkeldey2014 categorised uncertainty according to five criteria depicted in fig-tax which considers if a visualisation is implicit or explicit, intrinsic or extrinsic, visually integrable or seperable, coincident or adjacent, and static or dynamic. A similar version of this taxonomy was presented by @uncertchap2022 who commented that visualisation can be organised into two categories, "graphical annotations of distributional properties" and "visual encodings of uncertainty" which seems to functionally align with the intrinsic/extrinsic distinction by @Kinkeldey2014. @Griethe2006 organised uncertainty visualisations into two cases (1) a hypothesis test was preformed to confirm the validity of the visualisation and (2) the visualisation has uncertainty depicted. @Potter2012 organised several existing uncertainty visualisations into groups based on the dimensionality of the data (1D, 2D, 3D, and No Dimension) and the dimensionality of the uncertainty (Scalar, Vector, Tensor).  However, because the term "PDF", function that is used to describe a random variable, is used to describe both the data and the uncertainty for all dimensions. @Grewal2021 created a taxonomy that mapped uncertainty visualisations to some point in a 2D space defined by the "domain expertise" and "continuum of discreteness" (that scaled from "point estimate" to "continuous distribution").

>>>>

# Removed Sections
After all, if we consider a choropleth map, which ignores uncertainty entirely to be at one end of the spectrum (it is a map that only cares about showing signals reguardless of whether or not they are valid) then a completely colourless map is at the other end of the spectrum (it is a map that only cares about suppressing signals regardless of whether or not they are valid).

@Kinkeldey2014 categorised uncertainty according to five criteria depicted in fig-tax which considers if a visualisation is implicit or explicit, intrinsic or extrinsic, visually integrable or seperable, coincident or adjacent, and static or dynamic. A similar version of this taxonomy was presented by @uncertchap2022 who commented that visualisation can be organised into two categories, "graphical annotations of distributional properties" and "visual encodings of uncertainty" which seems to functionally align with the intrinsic/extrinsic distinction by @Kinkeldey2014. @Griethe2006 organised uncertainty visualisations into two cases (1) a hypothesis test was preformed to confirm the validity of the visualisation and (2) the visualisation has uncertainty depicted. @Potter2012 organised several existing uncertainty visualisations into groups based on the dimensionality of the data (1D, 2D, 3D, and No Dimension) and the dimensionality of the uncertainty (Scalar, Vector, Tensor).  However, because the term "PDF", function that is used to describe a random variable, is used to describe both the data and the uncertainty for all dimensions. @Grewal2021 created a taxonomy that mapped uncertainty visualisations to some point in a 2D space defined by the "domain expertise" and "continuum of discreteness" (that scaled from "point estimate" to "continuous distribution").
